\texttt{JAXbind}: Bind any function to JAX

### 2024-06-25 paper/paper.tex
Sync tex

JAX is widely used in machine learning and scientific computing, the
latter of which often relies on existing high-performance code that we
would ideally like to incorporate into JAX. Reimplementing the existing
code in JAX is often impractical and the existing interface in JAX for
binding custom code requires deep knowledge of JAX and its C++ backend.
The goal of \texttt{JAXbind} is to drastically reduce the effort
required to bind custom functions implemented in other programming
languages to JAX. Specifically, \texttt{JAXbind} provides an easy-to-use
Python interface for defining custom so-called JAX primitives that
support arbitrary JAX transformations.

@@

The use of JAX (Bradbury et al., 2018) is widespread in the natural
sciences. Of particular interest is JAX's powerful transformation
system. It enables to retrieve arbitrary derivatives of functions, batch
computations, and just-in-time compile code for additional performance.
Its transformation system relies on all components of the computation
being written in JAX.

A plethora of high-performance code is not written in JAX and thus not
accessible from within JAX. Rewriting these is often infeasible and/or
inefficient. Ideally, we would like to mix existing high-performance
code with JAX code. However, connecting code to JAX requires knowledge
of the internals of JAX and its C++ backend.

In this paper, we present \texttt{JAXbind}, a package for bridging any
function to JAX without in-depth knowledge of JAX's transformation
system. The interface is accessible from Python without requiring any
development in C++. The package is able to register any function, its
partial derivatives and their transpose functions as a JAX native call,
a so-called primitive.

@@

The functionality of \texttt{JAXbind} extends the external callback
functionality in JAX. Currently, \texttt{JAXbind}, akin to the external
callback functions in JAX, briefly requires python's global interpreter
lock (GIL) to call the user specified python function. In contrast to
JAX's external callback functions, it allows for both a custom
Jacobian-vector product and vector-Jacobian product. To the best of our
knowledge no other code currently exists for connecting generic
functions and both of their Jacobian products to JAX. The package that
comes the closest is Enzyme-JAX (EnzymeAD, 2024). Enzyme-JAX allows one
to differentiate a C++ function with Enzyme (W. Moses \& Churavy, 2020;
W. S. Moses et al., 2021, 2022) and connect it together with its
derivative to JAX. However, it enforces the use of Enzyme for deriving
derivatives and does not allow for connecting arbitrary code to JAX.

@@

Automatic differentiation is a core feature of JAX and often one of the
main reasons for using it. Thus, it is essential that custom functions
registered with JAX support automatic differentiation. In the following,
we will outline which functions our package respectively JAX requires to
enable automatic differentiation. For simplicity, we assume that we want
to connect the nonlinear function \(f(x_1,x_2) = x_1x_2^2\) to JAX. The
\texttt{JAXbind} package expects the Python function for \(f\) to take
three positional arguments. The first argument, \texttt{out}, is a
\texttt{tuple} into which the results are written. The second argument
is also a \texttt{tuple} containing the input to the function, in our
case, \(x_1\) and \(x_2\). Via \texttt{kwargs\_dump}, any keyword
arguments given to the registered JAX primitive can be forwarded to
\(f\) in a serialized form.

@@

To just-in-time compile the function, JAX needs to abstractly evaluate
the code, i.e.~it needs to be able to infer the shape and dtype of the
output of the function given only the shape and dtype of the input. We
have to provide these abstract evaluation functions returning the output
shape and dtype given an input shape and dtype for \(f\) as well as for
the \texttt{vjp} application. The output shape of the \texttt{jvp} is
identical to the output shape of \(f\) itself and does not need to be
specified again. The abstract evaluation functions take normal
positional and keyword arguments.

@@

In scientific computing, linear functions such as, e.g., spherical
harmonic transforms are widespread. If the function \(f\) is linear,
differentiation becomes trivial. Specifically for a linear function
\(f\), the pushforward respectively the \texttt{jvp} of \(f\) is
identical to \(f\) itself and independent of the position at which it is
computed. Expressed in formulas, \(\partial f(x)(dx) = f(dx)\) if \(f\)
is linear in \(x\). Analogously, the pullback respectively the
\texttt{vjp} becomes independent of the initial position and is given by
the linear transpose of \(f\), thus
\((\partial f(x))^{T}(dy) = f^T(dy)\). Also, all higher order
derivatives can be expressed in terms of \(f\) and its transpose. To
make use of these simplifications, \texttt{JAXbind} provides a special
interface for linear functions, supporting higher order derivatives,
only requiring an implementation of the function and its transpose.

@@

Currently, \texttt{JAXbind} only supports primitives that act on CPU
memory. In the future, GPU support could be added analogous to the CPU
backend. The automatic differentiation in JAX is backend agnostic and
would thus not require any additional bindings to work on the GPU.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Resolve2024}{}%
Arras, P., Roth, J., Ding, S., Reinecke, M., Fuchs, R., \& Johnson, V.
(2024). \emph{RESOLVE}. \url{https://gitlab.mpcdf.mpg.de/ift/resolve}

@@

\leavevmode\hypertarget{ref-Edenhofer2023NIFTyRE}{}%
Edenhofer, G., Frank, P., Roth, J., Leike, R. H., Guerdi, M.,
Scheel-Platz, L. I., Guardiani, M., Eberle, V., Westerkamp, M., \&
Enßlin, T. A. (2024). \emph{{Re-Envisioning Numerical Information Field
Theory (NIFTy.re): A Library for Gaussian Processes and Variational
Inference}}. \url{https://doi.org/10.48550/arXiv.2402.16683}

\leavevmode\hypertarget{ref-Moses2024}{}%
EnzymeAD. (2024). \emph{{Enzyme-JAX}} (Version 0.0.6) {[}Computer
software{]}. \url{https://github.com/EnzymeAD/Enzyme-JAX}

@@

>>>>

JAX is widely used in machine learning and scientific computing, the
latter of which often relies on existing high-performance code that we
would ideally like to incorporate into JAX. Reimplementing the existing
code in JAX is often impractical and the existing interface in JAX for
binding custom code either limits the user to a single Jacobian product
or requires deep knowledge of JAX and its C++ backend for general
Jacobian products. With \texttt{JAXbind} we drastically reduce the
effort required to bind custom functions implemented in other
programming languages with full support for Jacobian-vector products and
vector-Jacobian products to JAX. Specifically, \texttt{JAXbind} provides
an easy-to-use Python interface for defining custom, so-called JAX
primitives. Via \texttt{JAXbind}, any function callable from Python can
be exposed as a JAX primitive. \texttt{JAXbind} allows a user to
interface the JAX function transformation engine with custom derivatives
and batching rules, enabling all JAX transformations for the custom
primitive.

@@

The use of JAX (Bradbury et al., 2018) is widespread in the natural
sciences. Of particular interest is JAX's powerful transformation
system. It enables a user to retrieve arbitrary derivatives of
functions, batch computations, and just-in-time compile code for
additional performance. Its transformation system requires that all
components of the computation are written in JAX.

A plethora of high-performance code is not written in JAX and thus not
accessible from within JAX. Rewriting these codes is often infeasible
and/or inefficient. Ideally, we would like to mix existing
high-performance code with JAX code. However, connecting code to JAX
requires knowledge of the internals of JAX and its C++ backend.

In this paper, we present \texttt{JAXbind}, a package for bridging any
function to JAX without in-depth knowledge of JAX's transformation
system. The interface is accessible from Python without requiring any
development in C++. The package is able to register any function and its
partial derivatives and their transpose functions as a JAX native call,
a so-called primitive.

@@

The functionality of \texttt{JAXbind} extends the external callback
functionality in JAX. Currently, \texttt{JAXbind}, akin to the external
callback functions in JAX, briefly requires Python's global interpreter
lock (GIL) to call the user-specified Python function. In contrast to
JAX's external callback functions, \texttt{JAXbind} allows for both a
custom Jacobian-vector product and vector-Jacobian product. To the best
of our knowledge no other code currently exists for easily binding
generic functions and both of their Jacobian products to JAX, without
the need for C++ or LLVM. The package that comes the closest is
Enzyme-JAX (W. S. Moses \& Zinenko, 2024), which allows one to bind
arbitrary LLVM/MLIR, including C++, with automatically-generated (W.
Moses \& Churavy, 2020; W. S. Moses et al., 2021, 2022) or
manually-defined derivatives to JAX.

PyTorch (Ansel et al., 2024) and TensorFlow (Abadi et al., 2015) also
provide interfaces for custom extensions. PyTorch has an extensively
documented Python interface\footnote{\url{https://pytorch.org/docs/stable/notes/extending.html}}
for wrapping custom Python functions as PyTorch functions. This
interface connects the custom function to PyTorch's automatic
differentiation engine, allowing for custom Jacobian and Jacobian
transposed applications, similar to what is possible with JAXbind.
Additionally, PyTorch allows a user to interface its C++ backend with
custom C++ or CUDA extensions\footnote{\url{https://pytorch.org/tutorials/advanced/cpp_extension.html}}.
JAXbind, in contrast, currently only supports functions executed on the
CPU, although the JAX built-in C++ interface also allows for custom GPU
kernels. TensorFlow includes a C++ interface\footnote{\url{https://www.tensorflow.org/guide/create_op}}
for custom functions that can be executed on the CPU or GPU. Custom
gradients can be added to these functions.

@@

Automatic differentiation is a core feature of JAX and often one of the
main reasons for using it. Thus, it is essential that custom functions
registered with JAX support automatic differentiation. In the following,
we will outline which functions our package requires to enable automatic
differentiation via JAX. For simplicity, we assume that we want to
connect the nonlinear function \(f(x_1,x_2) = x_1x_2^2\) to JAX. The
\texttt{JAXbind} package expects the Python function for \(f\) to take
three positional arguments. The first argument, \texttt{out}, is a
\texttt{tuple} into which the results are written. The second argument
is also a \texttt{tuple} containing the input to the function, in our
case, \(x_1\) and \(x_2\). Via \texttt{kwargs\_dump}, any keyword
arguments given to the registered JAX primitive can be forwarded to
\(f\) in a serialized form.

@@

To just-in-time compile the function, JAX needs to abstractly evaluate
the code, i.e., it needs to be able to infer the shape and dtype of the
output of the function given only the shape and dtype of the input. We
have to provide these abstract evaluation functions returning the output
shape and dtype given an input shape and dtype for \(f\) as well as for
the \texttt{vjp} application. The output shape of the \texttt{jvp} is
identical to the output shape of \(f\) itself and does not need to be
specified again. The abstract evaluation functions take normal
positional and keyword arguments.

@@

In scientific computing, linear functions such as, e.g., spherical
harmonic transforms are widespread. If the function \(f\) is linear,
differentiation becomes trivial. Specifically for a linear function
\(f\), the pushforward or \texttt{jvp} of \(f\) is identical to \(f\)
itself and independent of the position at which it is computed.
Expressed in formulas, \(\partial f(x)(dx) = f(dx)\) if \(f\) is linear
in \(x\). Analogously, the pullback or \texttt{vjp} becomes independent
of the initial position and is given by the linear transpose of \(f\),
thus \((\partial f(x))^{T}(dy) = f^T(dy)\). Also, all higher order
derivatives can be expressed in terms of \(f\) and its transpose. To
make use of these simplifications, \texttt{JAXbind} provides a special
interface for linear functions, supporting higher order derivatives,
only requiring an implementation of the function and its transpose.

@@

Currently, \texttt{JAXbind} only supports primitives that act on CPU
memory. In the future, GPU support could be added, which should work
analogously to the CPU support in most respects. The automatic
differentiation in JAX is backend agnostic and would thus not require
any additional bindings to work on the GPU.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-tensorflow2015}{}%
Abadi, M., Agarwal, A., Barham, P., Brevdo, E., Chen, Z., Citro, C.,
Corrado, G. S., Davis, A., Dean, J., Devin, M., Ghemawat, S.,
Goodfellow, I., Harp, A., Irving, G., Isard, M., Jia, Y., Jozefowicz,
R., Kaiser, L., Kudlur, M., \ldots{} Zheng, X. (2015).
\emph{{TensorFlow}: Large-scale machine learning on heterogeneous
systems}. \url{https://www.tensorflow.org/}

\leavevmode\hypertarget{ref-PyTorch2024}{}%
Ansel, J., Yang, E., He, H., Gimelshein, N., Jain, A., Voznesensky, M.,
Bao, B., Bell, P., Berard, D., Burovski, E., Chauhan, G., Chourdia, A.,
Constable, W., Desmaison, A., DeVito, Z., Ellison, E., Feng, W., Gong,
J., Gschwind, M., \ldots{} Chintala, S. (2024, April). {PyTorch} 2:
Faster machine learning through dynamic {P}ython bytecode transformation
and graph compilation. \emph{29th ACM International Conference on
Architectural Support for Programming Languages and Operating Systems,
Volume 2 (ASPLOS '24)}. \url{https://doi.org/10.1145/3620665.3640366}

\leavevmode\hypertarget{ref-Resolve2024}{}%
Arras, P., Roth, J., Ding, S., Reinecke, M., Fuchs, R., \& Johnson, V.
(2024). \emph{RESOLVE}. \url{https://gitlab.mpcdf.mpg.de/ift/resolve}

@@

\leavevmode\hypertarget{ref-Edenhofer2023NIFTyRE}{}%
Edenhofer, G., Frank, P., Roth, J., Leike, R. H., Guerdi, M.,
Scheel-Platz, L. I., Guardiani, M., Eberle, V., Westerkamp, M., \&
Enßlin, T. A. (2024). Re-envisioning numerical information field theory
({NIFTy.re}): A library for {G}aussian processes and variational
inference. \emph{Journal of Open Source Software}, \emph{9}(98), 6593.
\url{https://doi.org/10.21105/joss.06593}

@@

\leavevmode\hypertarget{ref-Moses2024}{}%
Moses, W. S., \& Zinenko, O. (2024). \emph{{Enzyme-JAX}} (Version 0.0.6)
{[}Computer software{]}. \url{https://github.com/EnzymeAD/Enzyme-JAX}
